Two subtle problems with over-representation analysis

### 2024-09-16 manuscript/manuscript.Rmd
updates for reviewer comments

According to a study that used simulated differential expression data, FCS accuracy was
superior to ORA over a range of experimental designs [@Kaspi2020-rn].
Despite this performance difference, a survey of articles published in 2019 showed that
seven of the top eight most popular tools were based on ORA [@Wijesooriya2022-fw].
The discrepancy between performance and popularity is likely due to the relative ease of
conducting ORA as compared to FCS.
A minimal ORA might involve pasting a list of gene symbols into a text box on a website,
with results appearing nearly instantly.
On the other hand, FCS typically involves installing specific software and dealing with a
dataset representing every detected gene (~20,000 rows) and ensuring that the file
format from upstream tools is compatible with FCS packages.

@@

To quantify the effect of these two problems on RNA-seq based enrichment analysis, data from seven
transcriptomic studies were downloaded from DEE2 using the getDEE2 R/Bioconductor package
[@Ziemann2019-ux].
Raw data for these seven studies are available from NCBI Sequence Read Archive under the
accession numbers SRP128998, SRP038101, SRP037718, SRP096177, SRP097759, SRP253951 and
SRP068733 [@Felisbino2021-hc; @Lund2014-ds; @Rafehi2014-rv; @Keating2014-ab;
@Blanco-Melo2020-hh; @Rafehi2017-qn].
For each dataset, kallisto (v0.43.1) [@Bray2016-lu] transcript counts were aggregated to the
gene level.
Genes with fewer than 10 reads per sample on average were removed from downstream analysis.
Remaining genes passing this selection were included in the background gene set.
Differential expression analysis was conducted with DESeq2 v1.44.0 [@Love2014-wg], with
genes identified by their Ensembl identifiers.
Gene symbols were then fetched using biomaRt v2.60.0, based on Ensembl version 112 [@Durinck2009-dr].

@@

RNA-seq counts for SRA run accession ERR2539161 were obtained from DEE2 [@Mo2018-zb].
This dataset was selected due to its high number of assigned reads (367M).
A library of 200 random gene sets were generated, each containing 30 genes.
To generate differential expression profiles, 5% of gene sets were selected to be
differentially expressed, with equal numbers of up- and down-regulated pathways.
Pseudosamples were generated by first downsampling the counts using the `thincounts`
function of the `edgeR` package v4.2.0 [@Robinson2010-eb] to 20M reads, followed by the
addition of some extra variation drawn from a normal distribution with mean of 1 and
standard deviation varying between 0.1 and 0.6, this value selected randomly for each gene.
Three "control" and five "case" pseudosamples were generated.
For the genes selected to be differentially expressed, expression values were multiplied by a
log fold change in the case samples of 0.5 for upregulated genes and -0.5 for downregulated
genes.
Genes that were selected to be both up and downregulated were left unchanged.
With these expression profiles, DESeq2 was used for differential analysis, and the results
were subjected to clusterProfiler which suffers from both problems,
clusterProfiler with the background problem fix, clusterProfiler with the FDR problem fix,
clusterProfiler with the fixes to background and FDR problems, an ORA function called `fora`
from `fgsea` bioconductor package v1.30.0 that does not suffer these problems, and `fgsea` a
fast implementation of the preranked gene set enrichment analysis (GSEA) algorithm
[@Korotkevich2016-gd].
For the ORA functions above, significantly differentially expressed genes with FDR 0.05 were
selected for the foreground gene set.
If fewer than 200 genes were detected as significant, then 200 genes with the smallest
p-values were used for ORA, like a previous report [@Tarca2013-xh].
For `fgsea`, the DESeq2 test statistic was used for scoring differential expression.
After enrichment analysis with these approaches, the significant up and down regulated
sets were selected and compared to the ground truth to calculate precision, recall
and F1 score.
At each value of added variance, 1000 replications were conducted.
No set seed was used, but results were similar when run on different computer systems.
All analyses were undertaken with R v4.4.0 in a Docker container.

@@

ClusterProfiler is not the only tool to suffer from these problems, they are widespread
(**Table 1**). Still, there are some tools free of such problems.
For example, fora is a good option for R-based workflows, however the function does not calculate
the fold enrichment score,
so users will need to do it themselves [@Korotkevich2016-gd].
ShinyGO is an excellent alternative for web based analysis [@Ge2020-lu].
It doesn't suffer from the two issues described here, it is easy to use,
and has a high degree of reproducibility thanks to its stand-alone versioned docker
container option for local execution.
One problem that all of these tools listed in Table 1 have is that a background list is optional.
Sometimes this feature is hidden under "advanced options".
Given the strong biological and technical biases that are present in contemporary omics like single cell and spatial transcriptomics,
it should be a mandatory step for users to provide a custom background list.

@@

: Table 1. Classification of several popular freely available ORA tools regarding the provision of information such as
enrichment scores and FDR values, as well as behaviours regarding the two algorithmic issues described above.
(*) Denotes functions where options are provided to adjust this behaviour and are documented in the user manual.
(**) Documentation states unannotated genes are discarded with default settings using whole genome
background, but not when a custom background is used.
(+) Denotes that although no FDR values are provided, the results include sets where no overlaps were found, so users can adjust raw p-values themselves.
(^) Denotes enrichment scores will be included in the next Bioconductor release.

The purpose of this work was to thoroughly characterise two subtle problems with ORA, but there
are some bigger problems noted.
Firstly, all variations of ORA were less accurate (based on F1 score) as compared to FCS in this
simulation work, confirming a previous report [@Kaspi2020-rn].
Specifically, precision of ORA deteriorated with greater levels of added noise, while FCS
was relatively stable.
This means researchers can be more confident with FCS results, even if the data are
noisy.
The higher recall of FCS demonstrated in simulated differential expression profiles and confirmed
with downsampled real cancer RNA-seq data means that researchers will have more power to identify
pathways relevant to the biological processes being studied.
Secondly, many popular ORA tools listed in Table 1 do not provide enrichment scores in their
default output tables.
Enrichment scores are a surrogate for effect size, and without them, users may over-rely
on significance values when interpreting enrichment results.
Lastly, ORA results can vary dramatically depending on the number of genes selected in the
foreground, a mostly arbitrary choice.

>>>>

According to a study that used simulated differential expression data, FCS accuracy was
superior to ORA over a range of experimental designs [@Kaspi2020-rn].
A study of 1366 PubMed Central articles featuring enrichment analysis published in 2019
indicates that seven of the top eight most popular tools were based on ORA (according to the
supplementary data)[@Wijesooriya2022-fw].
The discrepancy between performance and popularity is likely due to the relative ease of
conducting ORA as compared to FCS.
A minimal ORA might involve pasting a list of gene symbols into a text box on a website,
with results appearing nearly instantly.
On the other hand, FCS typically involves installing specific software and dealing with a
dataset representing every detected gene (~20,000 rows) and ensuring that the file
format from upstream tools is compatible with FCS packages.

@@

To quantify the effect of these two problems on RNA-seq based enrichment analysis, data from seven
transcriptomic studies were downloaded from DEE2 using the getDEE2 R/Bioconductor package
[@Ziemann2019-ux].
The criteria for selection included human as the species, presence in the DEE2 database,
passing most DEE2 quality control measures, a simple control-case experiment design with
two or three replicates per group, and with 50 or more differentially expressed genes
(FDR<0.05).
Raw data for these seven studies are available from NCBI Sequence Read Archive under the
accession numbers SRP128998, SRP038101, SRP037718, SRP096177, SRP097759, SRP253951 and
SRP068733 [@Felisbino2021-hc; @Lund2014-ds; @Rafehi2014-rv; @Keating2014-ab;
@Blanco-Melo2020-hh; @Rafehi2017-qn].
For each dataset, kallisto (v0.43.1) [@Bray2016-lu] transcript counts were aggregated to the
gene level.
Genes with fewer than 10 reads per sample on average were removed from downstream analysis.
Remaining genes passing this selection were included in the background gene set.
Differential expression analysis was conducted with DESeq2 v1.44.0 [@Love2014-wg], with
genes identified by their Ensembl identifiers.
Gene symbols were then fetched using biomaRt v2.60.0, based on Ensembl version 112 [@Durinck2009-dr].

| SRA accession | Control accessions | Case accessions | Comparison | No. detected genes | No. genes with FDR<0.05 | Reference |
| --- | --- | --- | --- | --- | --- | --- |
| SRP128998 | SRR6467485, SRR6467486, SRR6467487 | SRR6467479, SRR6467480, SRR6467481 | Normal glucose versus high glucose | 15,635 | 3,472  | [@Felisbino2021-hc] |
| SRP038101 | SRR1171523, SRR1171524, SRR1171525 | SRR1171526, SRR1171527, SRR1171528 | Control versus azacytidine treated cells | 13,926 | 3,590 | [@Lund2014-ds] |
| SRP037718 | SRR1168228, SRR1168229, SRR1168230 | SRR1168225, SRR1168226, SRR1168227 | Control versus SAHA treated cells | 15,477 | 4,988 | [@Rafehi2014-rv] |
| SRP096177 | SRR5150595, SRR5150596, SRR5150597 | SRR5150592, SRR5150593, SRR5150594 | Control versus Set7 knock-down cells | 15,607 | 5,152 | [@Keating2014-ab] |
| SRP097759 | SRR5201525, SRR5201526 | SRR5201527, SRR5201528 | GFP control vs SAHH overexpression | 19,139 | 62 | Unpublished |
| SRP253951 | SRR11517674, SRR11517675, SRR11517676 | SRR11517677, SRR11517678, SRR11517679 | Mock vs SARS-CoV-2 infection | 15,182 | 8,588 | [@Blanco-Melo2020-hh] |
| SRP068733 | SRR3112216, SRR3112217, SRR3112218 | SRR3112219, SRR3112220, SRR3112221 | Control vs EP300 knock-down | 14,255 | 7,365 | [@Rafehi2017-qn] |

: Table 1. Seven selected RNA-seq datasets for benchmarking.

@@

RNA-seq counts for SRA run accession ERR2539161 were obtained from DEE2 [@Mo2018-zb].
This dataset was selected due to its high number of assigned reads (367M).
A library of 200 random gene sets, each containing 30 genes, was generated.
To generate differential expression profiles, 5% of gene sets were selected to be
differentially expressed, with equal numbers of up- and down-regulated pathways.
Pseudosamples were generated by first downsampling the counts using the `thincounts`
function of the `edgeR` package v4.2.0 [@Robinson2010-eb] to 20M reads, followed by the
addition of some extra variation drawn from a normal distribution with mean of 1 and
standard deviation varying between 0.1 and 0.6, this value selected randomly for each gene.
Three "control" and five "case" pseudosamples were generated.
For the genes selected to be differentially expressed, expression values were multiplied by a
log fold change in the case samples of 0.5 for upregulated genes and -0.5 for downregulated
genes.
Genes that were selected to be both up and downregulated were left unchanged.
With these expression profiles, DESeq2 was used for differential analysis, and the results
were subjected to clusterProfiler which suffers from both problems,
clusterProfiler with the background problem fix, clusterProfiler with the FDR problem fix,
clusterProfiler with the fixes to background and FDR problems, an ORA function called `fora`
from `fgsea` bioconductor package v1.30.0 that does not suffer these problems, and `fgsea` a
fast implementation of the preranked gene set enrichment analysis (GSEA) algorithm
[@Korotkevich2016-gd].
For the ORA functions above, significantly differentially expressed genes with FDR 0.05 were
selected for the foreground gene set.
If fewer than 200 genes were detected as significant, then 200 genes with the smallest
p-values were used for ORA, like a previous report [@Tarca2013-xh].
For `fgsea`, the DESeq2 test statistic was used for scoring differential expression.
After enrichment analysis with these approaches, the significant up and down regulated
sets were selected and compared to the ground truth to calculate precision, recall
and F1 score.
At each value of added variance, 1000 replications were conducted.
No set seed was used, but results were similar when run on different computer systems.
All analyses were undertaken with R v4.4.0 in a Docker container.

@@

ClusterProfiler is not the only tool to suffer from these problems, they are widespread
(**Table 2**). Still, there are some tools free of such problems.
For example, fora is a good option for R-based workflows, however the function does not calculate
the fold enrichment score,
so users will need to do it themselves [@Korotkevich2016-gd].
ShinyGO is an excellent alternative for web based analysis [@Ge2020-lu].
It doesn't suffer from the two issues described here, it is easy to use,
and has a high degree of reproducibility thanks to its stand-alone versioned docker
container option for local execution.
One problem that all of these tools listed in Table 2 have is that a background list is optional.
Sometimes this feature is hidden under "advanced options".
Given the strong biological and technical biases that are present in contemporary omics like single cell and spatial transcriptomics,
it should be a mandatory step for users to provide a custom background list.

@@

: Table 2. Classification of several popular freely available ORA tools regarding the provision of information such as
enrichment scores and FDR values, as well as behaviours regarding the two algorithmic issues described above.
(*) Denotes functions where options are provided to adjust this behaviour and are documented in the user manual.
(**) Documentation states unannotated genes are discarded with default settings using whole genome
background, but not when a custom background is used.
(+) Denotes that although no FDR values are provided, the results include sets where no overlaps were found, so users can adjust raw p-values themselves.
(^) Denotes enrichment scores will be included in the next Bioconductor release.

The purpose of this work was to thoroughly characterise two subtle problems with ORA, but there
are some bigger problems noted.
Firstly, all variations of ORA were less accurate (based on F1 score) as compared to FCS in this
simulation work, confirming a previous report [@Kaspi2020-rn].
Specifically, precision of ORA deteriorated with greater levels of added noise, while FCS
was relatively stable.
This means researchers can be more confident with FCS results, even if the data are
noisy.
The higher recall of FCS demonstrated in simulated differential expression profiles and confirmed
with downsampled real cancer RNA-seq data means that researchers will have more power to identify
pathways relevant to the biological processes being studied.
Secondly, many popular ORA tools listed in Table 2 do not provide enrichment scores in their
default output tables.
Enrichment scores are a surrogate for effect size, and without them, users may over-rely
on significance values when interpreting enrichment results.
Lastly, ORA results can vary dramatically depending on the number of genes selected in the
foreground, a mostly arbitrary choice.
